Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN and SVA normlaisation)

### 2017-08-07 code/Rmd/HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd
Added markdown to process HBCC data with surrogate variable analysis

---
title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN normalisation (DLPFC)';

thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN.Rmd'

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=5, fig.width=6, results='asis', cache = FALSE}
indToRemove = c("CMC_HBCC_RNA_PFC_3402")

@@

### Sample clustering
```{r decompse.normalise.data, fig.height=5, fig.width=5, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Gender, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=5, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Gender", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
residualSigCovars = runPCAandPlotCorrelations(tmp, COVARIATES,
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```

### Sanity check
```{r residualSigCovars.manual, fig.width=12, fig.height=8}
residualSigCovars[["PC_res"]][[2]]$plotData
```

### Residual calculation
Calculate weighted residuals and add back "Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION +
  ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

### Coexpression of residual expression 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
tmp = RESIDUAL.GENE_EXPRESSION
tmp[is.na(tmp)] = 0
cr = cor(t(tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual data
```{r decompse.normalise.data2, fig.height=5, fig.width=5, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(tmp, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Gender, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~., scales = 'free_y')
p
```

```{r decompse.normalise.data2.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue','Gender')])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(tmp)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue','Gender')]))
```

@@

### Differential expression analysis (SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=8, fig.width=8}
# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueBP_DLPFC-Dx.TissueControl_DLPFC",
                                     "Dx.TissueSCZ_DLPFC-Dx.TissueBP_DLPFC"),
                         levels = colnames(ADJUSTED.FIT$coefficients))
FIT.CONTR = contrasts.fit(ADJUSTED.FIT, contrasts=contrast)
FIT.CONTR = eBayes(FIT.CONTR)

# Get differnetial expression
DE = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR) 
names(DE) = gsub('Dx.Tissue', '', colnames(contrast))

DE = DE %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue','',Comparison),
                Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
  tidyr::separate(to.state, into = c('to.Dx','to.Tissue'), sep = '_') %>%
  tidyr::separate(ref.state, into = c('ref.Dx','ref.Tissue'), sep = '_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
DE$Direction[DE$adj.P.Val > 0.05 | abs(DE$logFC) < log2(1.2)] = 'NONE'

tmp = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2), ref.Dx != 'Other', to.Dx != 'Other') %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, ref.Dx, ref.Tissue, logFC) %>%
  group_by(to.Dx, to.Tissue, ref.Dx, ref.Tissue) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

kable(full_join(tmp,tmp1))

p = DE %>%
  dplyr::filter(ref.Dx != 'ALL', to.Dx != 'ALL') %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Dx+ref.Tissue~.+to.Dx+to.Tissue, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p
```

### Associate differential expression results with gc content, gene length and average expression
```{r associate.de, fig.height= 6, fig.width = 12}
pl = list()
pl[[1]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
                 aes(x = log10(Length), y = logFC, color = Direction)) + geom_point() 
pl[[1]] = pl[[1]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = log10(Length), y = logFC))
pl[[1]] = pl[[1]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[1]] = pl[[1]] + theme(legend.position = 'top')

pl[[2]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
                 aes(x = percentage_gc_content, y = logFC, color = Direction)) + geom_point()
pl[[2]] = pl[[2]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = percentage_gc_content, y = logFC))
pl[[2]] = pl[[2]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[2]] = pl[[2]] + theme(legend.position = 'top')

pl[[3]] = ggplot(DE %>% dplyr::filter(ref.Dx == 'Control', to.Dx == 'SCZ'), 
                 aes(x = AveExpr, y = logFC, color = Direction)) + geom_point()
pl[[3]] = pl[[3]] + geom_smooth(method = 'loess', inherit.aes = FALSE, aes(x = AveExpr, y = logFC))
pl[[3]] = pl[[3]] + facet_grid(ref.Tissue~.) + scale_color_manual(values = c('green','grey','red'))
pl[[3]] = pl[[3]] + theme(legend.position = 'top')

multiplot(plotlist = pl, cols = 3)
```

### Differential expression analysis (Gender specific SCZ - Control)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp1, fig.height=6, fig.width=12}
# Covariates to adjust
COVARIATES$Dx.Tissue.Gender = factor(paste(COVARIATES$Dx.Tissue, COVARIATES$Gender, sep = '_'))
postAdjustCovars = c('Dx.Tissue.Gender', 'RIN', 'IntronicRate')

writeLines(paste('Using following covariates in the model:',
                 paste(postAdjustCovars, collapse=', ')))

# Post adjusted design matrix
DESIGN1 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = F)$design
DESIGN1 = DESIGN1[,linColumnFinder(DESIGN1)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION1 = limma::voom(NEW.COUNTS, design = DESIGN1)
VOOM.GENE_EXPRESSION1$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT1 = lmFit(VOOM.GENE_EXPRESSION1)

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.Tissue.GenderSCZ_DLPFC_Male-Dx.Tissue.GenderControl_DLPFC_Male",
                                     "Dx.Tissue.GenderSCZ_DLPFC_Female-Dx.Tissue.GenderControl_DLPFC_Female"),
                         levels = colnames(ADJUSTED.FIT1$coefficients))
FIT.CONTR1 = contrasts.fit(ADJUSTED.FIT1, contrasts=contrast)
FIT.CONTR1 = eBayes(FIT.CONTR1)

# Get differnetial expression
DE1 = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR1) 
names(DE1) = gsub('Dx.Tissue.Gender', '', colnames(contrast))

DE1 = DE1 %>% 
  data.table::rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Comparison = gsub('Dx.Tissue.Gender','',Comparison),
                Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  tidyr::separate(Comparison, into = c('to.state','ref.state'), sep = '-') %>%
  tidyr::separate(to.state, into = c('to.Dx', 'to.Tissue', 'to.Gender'), sep = '_') %>%
  tidyr::separate(ref.state, into = c('ref.Dx', 'ref.Tissue', 'ref.Gender'), sep = '_') %>%
  left_join(Ensemble2HGNC) %>%
  left_join(GENE.LEN %>% dplyr::mutate(Length = as.numeric(Length)))
DE1$Direction[DE1$adj.P.Val > 0.05 | abs(DE1$logFC) < log2(1.2)] = 'NONE'

tmp = DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE1 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(ensembl_gene_id, to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender, logFC) %>%
  group_by(to.Dx, to.Tissue, to.Gender, ref.Dx, ref.Tissue, ref.Gender) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

kable(full_join(tmp,tmp1))

p = DE1 %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(ref.Tissue~.+ref.Gender, scales = 'fixed')
p = p + ggtitle('Differential Expression between Schizophrenia and Control')
p

table(COVARIATES$Dx.Tissue, COVARIATES$Gender)
```

### Differential expression analysis (ACC - DLPFC, SCZ - Control, Female - Male)
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp2, fig.height=10, fig.width=6}
postAdjustCovars = c('Dx.Tissue', 'Gender', 'RIN', 'IntronicRate')

# Post adjusted design matrix
DESIGN2 = getDesignMatrix(droplevels(COVARIATES[,postAdjustCovars,drop=F]),Intercept = T)$design
DESIGN2 = DESIGN2[,linColumnFinder(DESIGN2)$indepCols]

# Estimate voom weights
VOOM.GENE_EXPRESSION2 = voom(NEW.COUNTS, design = DESIGN2)
VOOM.GENE_EXPRESSION2$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
ADJUSTED.FIT2 = lmFit(VOOM.GENE_EXPRESSION2)

# Fit contrast
contrast = makeContrasts(contrasts=c("Dx.TissueSCZ_DLPFC-Dx.TissueControl_DLPFC",
                                     "GenderMale"),
                         levels = colnames(ADJUSTED.FIT2$coefficients))
FIT.CONTR2 = contrasts.fit(ADJUSTED.FIT2, contrasts=contrast)
FIT.CONTR2 = eBayes(FIT.CONTR2)

# Get differnetial expression
DE2 = lapply(1:dim(contrast)[2], function(i, FIT){
  topTable(FIT, coef=i, number = 50000, confint = TRUE) %>%
    rownameToFirstColumn('ensembl_gene_id') 
}, FIT.CONTR2) 
names(DE2) = c('SCZ-CONTROL', 'FEMALE-MALE')

DE2 = DE2 %>% 
  rbindlist(idcol = 'Comparison') %>%
  dplyr::mutate(Direction = sign(logFC),
                Direction = factor(Direction, c(-1,1), c('-1' = 'DOWN', '1' = 'UP')),
                Direction = as.character(Direction)) %>%
  left_join(Ensemble2HGNC) 
DE2$Direction[DE2$adj.P.Val > 0.05 | abs(DE2$logFC) < log2(1.2)] = 'NONE'

tmp = DE2 %>%
  dplyr::filter(adj.P.Val <= 0.05) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_UP = length(unique(ensembl_gene_id[logFC > 0])))

tmp1 = DE2 %>%
  dplyr::filter(adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) %>%
  dplyr::select(Comparison, ensembl_gene_id, logFC) %>%
  group_by(Comparison) %>%
  dplyr::summarise(FDR_0_05_FC_1.2_DOWN = length(unique(ensembl_gene_id[logFC < 0])),
                   FDR_0_05_FC_1.2_UP = length(unique(ensembl_gene_id[logFC > 0])))

kable(full_join(tmp,tmp1))

p = DE2 %>%
  ggplot(aes(y = -log10(adj.P.Val), x = logFC, color = Direction)) + geom_point() + xlim(c(-1,1))
p = p + scale_color_manual(values = c('green','grey','red')) + theme(legend.position = 'top')
p = p + facet_grid(Comparison~., scales = 'free_y')
p = p + ggtitle('Differential Expression')
p
```

### Gene set enrichment analysis
Using Fisher's exact test for gene set enrichment analysis
```{r enrich.de, fig.height=16, fig.width=16}
all.diff.exp = list(DE %>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE1%>%
                      tidyr::unite(Comparison1, to.Dx, to.Tissue, to.Gender) %>%
                      tidyr::unite(Comparison2, ref.Dx, ref.Tissue, ref.Gender) %>%
                      tidyr::unite(Comparison, Comparison1, Comparison2, sep = '-'), 
                    DE2) %>%
  data.table::rbindlist(fill = TRUE, use.names = TRUE)
diffexp.genes.up = all.diff.exp %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC > 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.up) = paste(names(diffexp.genes.up), 'UP', sep = '.')

diffexp.genes.down = all.diff.exp %>%
  dlply(.(Comparison), .fun = function(x){
    tmp = x$hgnc_symbol[x$adj.P.Val <= 0.05 & x$logFC < 0] %>% unique
    tmp = tmp[!is.na(tmp)]
  })
names(diffexp.genes.down) = paste(names(diffexp.genes.down), 'DOWN', sep = '.')

diffexp.genes = c(diffexp.genes.up, diffexp.genes.down)
diffexp.genes = diffexp.genes[sapply(diffexp.genes, length) > 20]
backgroundGenes = unique(DE$hgnc_symbol)

# Download all related genesets from synapse
geneset.files = synQuery('select * from file where parentId == "syn3240583"')
ALL_USED_IDs = c(ALL_USED_IDs, geneset.files$file.id)
genesets.to.test = sapply(geneset.files$file.name, function(dataSource, geneset.files){
  tmp = downloadFile(geneset.files$file.id[geneset.files$file.name == dataSource]) %>%
    dlply(.(Name), .fun = function(x){
      str_split(x$symBeforeOverlap, '\\|') %>% unlist %>% unique
    })
  }, geneset.files) %>%
  CovariateAnalysis::filterGeneSets(backgroundGenes) %>%
  do.call(c,.)

# Perform enrichment analysis
enrichment.results = llply(diffexp.genes, .fun = function(x, genesetsToTest, backgroundGenes){
   ldply(genesetsToTest, .fun = function(y,x,backgroundGenes){
     CovariateAnalysis::fisherEnrichment(x, y, backgroundGenes)
    }, x, backgroundGenes, .parallel = TRUE) %>%
    dplyr::mutate(fdr = p.adjust(pval, method = 'fdr'))
}, genesets.to.test, backgroundGenes) %>%
  rbindlist(idcol = 'Comparison', use.names = T, fill = T) %>%
  dplyr::rename(Category = .id)
```

@@

  normalizationType	= 'CQN'
)

# Code
CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store residual gene expression for network analysis
RESIDUAL.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_netResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
nEXP_OBJ = File('CMC_ACC_DLPFC_netResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for network analysis)', 
                parentId = CODE$properties$id)
annotations(nEXP_OBJ) = annotations(CODE)
annotations(nEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
nEXP_OBJ = synStore(nEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store differential expression results
write.table(DE, file = 'CMC_ACC_DLPFC_DiffExpression.tsv', sep = '\t', row.names=F, quote=F)
DEXP_OBJ = File('CMC_ACC_DLPFC_DiffExpression.tsv', 
                name = 'Differential Expression Results (Dx.Tissue)', 
                parentId = CODE$properties$id)
annotations(DEXP_OBJ) = annotations(CODE)
annotations(DEXP_OBJ)$dataSubType = 'diffExp'
DEXP_OBJ = synStore(DEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store enrichment results
write.table(enrichment.results, file = 'CMC_ACC_DLPFC_EnrichResults.tsv', sep = '\t', row.names=F, quote=F)
ENRICH_OBJ = File('CMC_ACC_DLPFC_EnrichResults.tsv', 
                  name = 'Enrichment Analysis Results (Dx.Tissue)', 
                  parentId = CODE$properties$id)
annotations(ENRICH_OBJ) = annotations(CODE)
annotations(ENRICH_OBJ)$dataSubType = 'enrichResults'
ENRICH_OBJ = synStore(ENRICH_OBJ, activity = synGetActivity(COV_OBJ$properties$id))
# stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (estimated)                         |  `r NCOUNT_OBJ$properties$id`    |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Residual Expression (for network analysis) |  `r nEXP_OBJ$properties$id`      |
|  Differential Expression                    |  `r DEXP_OBJ$properties$id`      |
|  Enrichment Analysis                        |  `r ENRICH_OBJ$properties$id`    |

>>>>

---
title: "Covariate analysis of HBCC RNASeq data reprocessed with STAR-feature counts workflow with Ensembl v75 (DLPFC with CQN and SVA normlaisation)"
author: "Thanneer Perumal"
date: "`r date()`"
output: html_document
---
```{r knit2synapse, eval=FALSE}
library(synapseClient)
library(knit2synapse)

@@

knit2synapse::knitToFolderEntity(file = "./HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd",
                                 parentId ="syn7501880",
                                 entityName = 'HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA')
```

@@

```{r synapse.parameters, include=FALSE, cache=TRUE}
parentId = 'syn7501880';
activityName = 'Covariate adjustments';
activityDescription = 'Covariate analysis of STAR aligned feature counts with Ensembl v75 with CQN and SVA normalisation (DLPFC)';

thisFileName <- 'HBCC_DLPFC_STAR_ENSEMBL_v75_CQN_SVA.Rmd'

@@

### Outlier analysis
Detect outlier samples based on their expression (logCPM) pattern
```{r detect.outliers, fig.height=5, fig.width=5, results='asis', cache = FALSE}
indToRemove = c("CMC_HBCC_RNA_PFC_3402")

@@

### Sample clustering
```{r decompse.normalise.data, fig.height=6, fig.width=8, results='asis', cache=FALSE}
# Find principal components of expression to plot
tmp = CQN.GENE_EXPRESSION$E
tmp[is.na(tmp)] = 0
PC <- prcomp(tmp, scale.=T, center = T)

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=Gender, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=0)
p
```
Tree based clustering of samples
```{r decompse.normalise.data.1, fig.height=6, fig.width=10, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c("Gender", "Dx.Tissue")])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

@@

# Residuals after normalisation
RESIDUAL.GENE_EXPRESSION = residuals.MArrayLM(ADJUSTED.FIT, CQN.GENE_EXPRESSION$E)
```

### SVA Adjustments for eQTL analysis
Conditioned on primary variable (Dx.Tissue) and covariates estimate surrogate variables using SVA package
```{r sva}
# Get (null) design matrix
DESIGN = DM1$design
MODEL0 = DESIGN[,-grep('Dx.Tissue',colnames(DESIGN))] # Get null model by removing variable of interest
MODEL0 = MODEL0[,linColumnFinder(MODEL0)$indepCols]
MODEL1 = DESIGN
MODEL1 = MODEL1[,linColumnFinder(MODEL1)$indepCols]

# Source modified functions from sva package
source('../R/irwsva.build.R')
source('../R/f.pvalue.R')

# Compute actual variance of all principal components
tmp = svd(RESIDUAL.GENE_EXPRESSION)
actual.var = tmp$d^2/sum(tmp$d^2)

# Compute permuted variance of all principal components
permuted.var = foreach(i = 1:20, .combine = rbind, .packages = c('dplyr')) %dopar% {
  tmp.residual = t(apply(RESIDUAL.GENE_EXPRESSION, 1, sample, replace = FALSE))
  tmp = svd(tmp.residual)
  permuted.var = tmp$d^2/sum(tmp$d^2)
}
permuted.var = apply(permuted.var, 2, mean)
NUM.SV = sum(actual.var >= permuted.var[1])

# Plot variance components
var.comp = data.frame(component = 1:length(actual.var), 
                      residual = actual.var, 
                      permuted.residual = permuted.var) %>%
  tidyr::gather(data, value, -component)
p = ggplot(var.comp %>% filter(component <= round(NUM.SV*1.2)), aes(x = component, y = value, color = data)) + geom_point()
p = p + geom_hline(yintercept=permuted.var[1], linetype = 'dashed') + xlab('Singular Dimension Index')
p = p + geom_vline(xintercept=NUM.SV, linetype = 'dashed') + ylab('Fraction of Variance Explained')
p = p + theme(legend.position = c(0.8, 0.8), legend.title = element_blank())
p

# Estimate surrogate variables
SURR.VAR = irwsva.build(CQN.GENE_EXPRESSION$E, MODEL1, MODEL0, n.sv = NUM.SV, B = 30, tol = 1e-20)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])
rownames(SURR.VAR) = rownames(MODEL1)

# Estimate voom weights for dispersion control
VOOM.SVA.WEIGHTS = voom(NEW.COUNTS, 
                        design=cbind(MODEL1, SURR.VAR), plot=F)
VOOM.SVA.WEIGHTS$E = CQN.GENE_EXPRESSION$E

# Fit linear model using new weights and new design
VOOM.SVA.FIT = lmFit(VOOM.SVA.WEIGHTS)
```

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Residuals after normalisation
RESIDUAL.SVA.GENE_EXPRESSION = residuals.MArrayLM(VOOM.SVA.FIT,
                                                  CQN.GENE_EXPRESSION$E)

# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DESIGN), value = T)
RESIDUAL.SVA.GENE_EXPRESSION = RESIDUAL.SVA.GENE_EXPRESSION + 
  VOOM.SVA.FIT$coefficients[,varsToAddIn] %*% t(VOOM.SVA.FIT$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.SVA.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
Clustering residual data
```{r decompse.normalise.data2, fig.height=6, fig.width=6, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.SVA.GENE_EXPRESSION, scale.=T, center = T)

@@

plotdata <- left_join(plotdata, rownameToFirstColumn(COVARIATES, 'SampleID')) %>%
  tidyr::separate(Dx.Tissue, c('Dx','Tissue'), sep = '_') %>%
  dplyr::mutate(label = SampleID)
# plotdata$label[!(plotdata$SampleID %in% top.nsamples)] = ''

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color = Dx, shape=Tissue, size=Age_of_Death))
p <- p + theme_bw() + theme(legend.position="right") + facet_grid(Tissue~.)
# p <- p + geom_text(aes(label= label), size=4, hjust=0)
p
```

```{r decompse.normalise.data2.1, fig.height=8, fig.width=12, results='asis'}
# Eucledian tree based analysis
COVARIATES.tmp = data.matrix(COVARIATES[,c('Dx.Tissue'),drop=FALSE])
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0

tree = hclust(as.dist(t(RESIDUAL.SVA.GENE_EXPRESSION)))
cols = WGCNA::labels2colors(COVARIATES.tmp);

WGCNA::plotDendroAndColors(tree, 
                           colors = cols, 
                           dendroLabels = FALSE, 
                           abHeight = 0.80, 
                           main = "Sample dendrogram",
                           groupLabels = colnames(COVARIATES.tmp[,c('Dx.Tissue'),drop=FALSE]))
```

@@

@@

  normalizationType	= 'SVA'
)

# Code
CODE <- Folder(name = "HBCC - DLPFC - STAR - FEATURE COUNTS - ENSEMBL v75 - CQN - SVA", parentId = parentId)
annotations(CODE) = all.annotations
CODE <- synStore(CODE)

@@

# Store design matrix
SURR.VAR %>%
  rownameToFirstColumn('SampleID') %>%
  write.table(file = 'CMC_ACC_DLPFC_SurrVar.tsv', sep = '\t', row.names=F, quote=F)
SURR.VAR_OBJ = File('CMC_ACC_DLPFC_SurrVar.tsv', name = 'Surrogate Variables', parentId = CODE$properties$id)
annotations(SURR.VAR_OBJ) = annotations(CODE)
annotations(SURR.VAR_OBJ)$dataSubType = 'surrogateVariable'
SURR.VAR_OBJ = synStore(SURR.VAR_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# Store residual gene expression for network analysis
RESIDUAL.SVA.GENE_EXPRESSION %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_eqtlResidualExpression.tsv', sep = '\t', row.names=F, quote=F)
eEXP_OBJ = File('CMC_ACC_DLPFC_eqtlResidualExpression.tsv', 
                name = 'Normalised, covariates removed residual expression (for eQTL analysis)', 
                parentId = CODE$properties$id)
annotations(eEXP_OBJ) = annotations(CODE)
annotations(eEXP_OBJ)$dataSubType = 'residualGeneExpForNetAnlz'
eEXP_OBJ = synStore(eEXP_OBJ, activity = synGetActivity(COV_OBJ$properties$id))

# stopCluster(cl)
```
|  *Results*                                  |  *SynapseID*                     |
|  -----------------------------------------  |   ---------                      |
|  Covariates                                 |  `r COV_OBJ$properties$id`       |
|  Counts (raw)                               |  `r COUNT_OBJ$properties$id`     |
|  Counts (estimated)                         |  `r NCOUNT_OBJ$properties$id`    |
|  Counts (lcpm)                              |  `r LCOUNT_OBJ$properties$id`    |
|  Offset (for gene length and gc content)    |  `r OFFSET_OBJ$properties$id`    |
|  Design Matrix                              |  `r DM_OBJ$properties$id`        |
|  Surrogate Variables                        |  `r SURR.VAR_OBJ$properties$id`  |
|  Residual Expression (for eQTL analysis)    |  `r eEXP_OBJ$properties$id`      |
